---
title: "Loss_making_route_analysis"
author: "Rupikasiva"
date: "2026-03-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readr)
library(ggplot2)
library(knitr)
library(dplyr)
library(pROC)
library(caret)
library(naniar)
library(corrplot)

```

## Loading the datatset

```{r echo=FALSE}
airline_df <- read.csv("airline_route_profitability.csv")
head(airline_df)
summary(airline_df)
dim(airline_df)

#Checking missing Values
missing_value <- colSums(is.na(airline_df))
view(missing_value)

vis_miss(airline_df) +
  scale_fill_manual(
    values = c("lightblue", "darkblue"),
    labels = c("Present (99.7%)", "Missing (0.3%)")
  ) + 
  labs (title= "Missing Data Pattern", fill = "Data Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Removing missing values since they are relatively very low and does not affect the output variable
airline_df <- na.omit(airline_df)
colSums(is.na(airline_df))

#Histogram for Profit analysis
hist(airline_df$Profit, main="Histogram of Profit", xlab = "Profit", ylab = "Frequency")
hist(airline_df$Profit_Margin, main="Histogram of Profit Margin", xlab = "Profit", ylab = "Frequency")

#Proceeding with Profit Margin
airline_df$route_classification <- ifelse(airline_df$Profit_Margin < 0, "Loss",
                                          ifelse(airline_df$Profit_Margin < 5, "Marginal","Profitable"))
view(airline_df)

```

This is the profit and loss analysis for the routes
```{r echo=FALSE}
#Checking route counts under each category

table(airline_df$route_classification)

library(dplyr)
library(knitr)
library(kableExtra)

route_summary <- airline_df %>%
  count(route_classification) %>%
  arrange(desc(n)) %>%
  rename(
    "Route Classification" = route_classification,
    "Number of Routes" = n
  )

kable(route_summary, caption = "Distribution of Route Classification") %>%
  kable_styling(full_width = FALSE, position = "center")

#Listing top 10 loss making routes with profit
#airline_df[order(airline_df$Profit), ][1:10, ]

#Listing top 10 loss making routes with Profit Margin
lowest_profit_margin <- airline_df %>%
  arrange(Profit_Margin) %>%
  slice(1:10) %>%
  select(Route, Route_Category, Demand_Level, Load_Factor,
         Flight_Hours, Total_Revenue, Total_Cost,
         Profit, Profit_Margin)
colnames(lowest_profit_margin) <- c("Route", "Category", "Demand",
                            "Load Factor", "Flight Hours",
                            "Revenue", "Cost", "Profit", "Profit_Margin")
  
  
kable(lowest_profit_margin, caption = "Top 10 loss making routes with Profit Margin")
#loss making routes are mostly short haul routes and the demand level is medium, and season is normal, 


library(knitr)
library(dplyr)

lowest_profit_margin <- airline_df %>%
  arrange(Profit_Margin) %>%
  slice(1:10) %>%
  select(Route, Route_Category, Demand_Level, Load_Factor,
         Flight_Hours, Total_Revenue, Total_Cost,
         Profit, Profit_Margin)

colnames(lowest_profit_margin) <- c("Route", "Category", "Demand",
                                   "Load Factor", "Flight Hours",
                                   "Revenue", "Cost", 
                                   "Profit", "Profit Margin")

kable(lowest_profit_margin,
      caption = "Top 10 Loss Making Routes by Profit Margin",
      digits = 2) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("bordered")
  )




```
Checking which cost variable contributes to the loss and profit making routes


```{r echo=FALSE}
cost_df <- airline_df %>%
  filter(route_classification %in% c("Loss", "Profitable"))

# Checking which cost variable contributes to the loss and profit making routes
cost_summary <- cost_df %>%
  group_by(route_classification) %>%
  summarise(
    Fuel_Cost = mean(Fuel_Cost),
    Crew_Cost = mean(Crew_Cost),
    Maintenance_Cost = mean(Maintenance_Cost),
    Airport_Fees = mean(Airport_Fees),
    Catering_Cost = mean(Catering_Cost),
    Handling_Cost = mean(Handling_Cost),
    Overhead_Cost = mean(Overhead_Cost),
    Marketing_Cost = mean(Marketing_Cost)
  ) %>%
  pivot_longer(
    cols = -route_classification,
    names_to = "Cost_Category",
    values_to = "Average_Cost"
  )

ggplot(cost_summary, aes(x = Cost_Category, y = Average_Cost, fill = route_classification)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Costs: Loss vs Profitable Routes",
    x = "Cost Category",
    y = "Average Cost",
    fill = "Route Group"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Checking Cost as percentage of Total Revenue

cost_pctg_df <- cost_df %>%
  mutate(
    Fuel_pctg = Fuel_Cost / Total_Revenue * 100,
    Crew_pctg = Crew_Cost / Total_Revenue * 100,
    Maintenance_pctg = Maintenance_Cost / Total_Revenue * 100,
    Airport_pctg = Airport_Fees / Total_Revenue * 100,
    Catering_pctg = Catering_Cost / Total_Revenue * 100,
    Handling_pctg = Handling_Cost / Total_Revenue * 100,
    Overhead_pctg = Overhead_Cost / Total_Revenue * 100,
    Marketing_pctg = Marketing_Cost / Total_Revenue * 100
  ) %>%
  group_by(route_classification) %>%
  summarise(
    Fuel_pctg = mean(Fuel_pctg),
    Crew_pctg = mean(Crew_pctg),
    Maintenance_pctg = mean(Maintenance_pctg),
    Airport_pctg = mean(Airport_pctg),
    Catering_pctg = mean(Catering_pctg),
    Handling_pctg = mean(Handling_pctg),
    Overhead_pctg = mean(Overhead_pctg),
    Marketing_pctg = mean(Marketing_pctg)
  ) %>%
  pivot_longer(
    cols = -route_classification,
    names_to = "Cost_Category",
    values_to = "Average_Percent_Revenue"
  )

ggplot(cost_pctg_df, aes(x = Cost_Category, y = Average_Percent_Revenue, fill = route_classification)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Cost Components as Percentage of Revenue",
    x = "Cost Category",
    y = "Average Cost (Percentage of Total Revenue)",
    fill = "Route Group"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Heat map 

```{r}
# Step 1 - Select only numeric variables relevant to analysis
numeric_vars <- airline_df[, c("Profit_Margin", 
                                "Load_Factor", 
                                "Flight_Hours",
                                "Passengers",
                                "Aircraft_Capacity",
                                "Ticket_Revenue",
                                "Total_Revenue",
                                "Fuel_Cost",
                                "Total_Cost",
                                "Profit")]

# Step 2 - Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Step 3 - Plot the heatmap
corrplot(cor_matrix, 
         method = "color", 
         type = "upper",            
         addCoef.col = "black",     
         tl.col = "black",          
         tl.srt = 45,               
         number.cex = 0.7,          
         title = "Correlation Heatmap of Numeric Variables",
         mar = c(0,0,1,0)         
)
```





Checking different factors that affect the routes

```{r echo=FALSE}

route_colors <- c(
  "Loss" = "#08306B",        # dark blue
  "Marginal" = "#4292C6",    # medium blue
  "Profitable" = "#C6DBEF"   # light blue
)




#Ticket Revenue Vs Route Classification
ggplot(airline_df, aes(x = route_classification, y = Ticket_Revenue, fill = route_classification)) +
  geom_boxplot() +
  scale_fill_manual(values = route_colors) +
  labs(
    title = "Ticket Revenue by Route Classification",
    x = "Route Group",
    y = "Ticket Revenue"
  ) +
  theme_minimal()

#Demand level Vs Route Classification
ggplot(airline_df, aes(x = Demand_Level, fill = route_classification)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = route_colors) +
  labs(
    title = "Demand Level by Route Classification",
    x = "Demand level",
    y = "No of Routes",
    fill = "Route Classification"
  ) +
  theme_minimal()

#Season Vs Route Classification

ggplot(airline_df, aes(x = Season, fill = route_classification)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = route_colors) +
  labs(
    title = "Route Classification across Seasons",
    x = "Season",
    y = "Number of Routes",
    fill = "Route Classification"
  ) +
  theme_minimal()

```


```{r}

#Load Factor Vs Route Classification
loadfactor <- ggplot(airline_df, aes(x = route_classification, y = Load_Factor, fill = route_classification)) +
  geom_boxplot() +
  scale_fill_manual(values = route_colors) +
  labs(
    title = "Load Factor Distribution",
    x = "Route Group",
    y = "Load Factor"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Load_factor_plot.png", plot = loadfactor, width = 7.55, height = 7, units = "cm")
```

Few more features Vs Route Classification


```{r echo=FALSE}
#Flight Hours Vs Route classification
ggplot(airline_df, aes(x= route_classification, y = Flight_Hours, fill = route_classification))+
  geom_boxplot() +
  scale_fill_manual(values = route_colors) +
  labs(
    title = "Flight Hours by Route Classification",
    x = "Route Classification",
    y = "Flight Hours"
  ) +
  theme_minimal()


#route category Vs route classification
ggplot(airline_df, aes(x = Route_Category, fill = route_classification)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = route_colors) +
  labs(
    title = "Route Classification by Route Category",
    x = "Route Category",
    y = "Number of Routes",
    fill = "Route Classification"
  ) +
  theme_minimal()

#Aircraft Type Vs Route Classification

aircraft_type <- ggplot(airline_df, aes(x = Aircraft_Type, fill = route_classification)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = route_colors) +
  labs(
    title = "Route Classification by Aircraft Type",
    x = "Aircraft Type",
    y = "Number of Routes",
    fill = "Route Classification"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("aircraft_type.png", plot = aircraft_type, width = 9.33, height = 5.85, units = "cm")


#Flight Hours Vs Profit Margin

profit_margin <- ggplot(airline_df, aes(x = Flight_Hours, y = Profit_Margin, color = route_classification)) +
  geom_jitter(alpha = 0.4, width = 0.2, height = 0) +
  scale_color_manual(values = route_colors) +
  labs(
    title = "Profit Margin vs Flight Hours",
    x = "Flight Hours",
    y = "Profit Margin",
    color = "Route Classification"
  ) +
  theme_minimal()

ggsave("profit_margin_plot.png", plot = profit_margin , width = 9.35, height = 9, units = "cm")


```

Logistic regression for predicting the profit or loss route classification with the analysed factors

Model 1

```{r echo=FALSE}
#Defining the binary variable 0 - Profit 1 - Loss
airline_df$loss_factor <- ifelse(airline_df$route_classification == "Loss", 1, 0)

#Train and Test - Data splitting
set.seed(1)

train_ind <- sample(1:nrow(airline_df), size = nrow(airline_df)*0.8)

df_train <- airline_df[train_ind, ]
df_test <- airline_df[-train_ind, ]

#Using operational features as per EDA done earlier
logreg.fit <- glm(loss_factor ~ Load_Factor + Flight_Hours +
                   Route_Category + Aircraft_Type + Demand_Level +
                   Season,
                   data = df_train,
                   family = binomial)

summary(logreg.fit)

pred_prob <- predict(logreg.fit, newdata = df_test, type = "response")
head(pred_prob)
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
head(pred_class)

table(Predicted = pred_class, Actual = df_test$loss_factor)

confusionMatrix(as.factor(pred_class),
                as.factor(df_test$loss_factor),
                positive = "1")


roc_obj <- roc(df_test$loss_factor, pred_prob)
auc_value <- auc(roc_obj)

plot(roc_obj,
     col = "blue",
     main = "ROC Curve for Route Loss Prediction")

legend("bottomright",
       legend = paste("AUC =", round(auc_value,3)))

```
Model 2 - aircraft capacity and removing demand level, aircraft_type

```{r}
#Defining the binary variable 0 - Profit 1 - Loss
airline_df$loss_factor <- ifelse(airline_df$route_classification == "Loss", 1, 0)

#Train and Test - Data splitting
set.seed(1)

train_ind <- sample(1:nrow(airline_df), size = nrow(airline_df)*0.8)

df_train <- airline_df[train_ind, ]
df_test <- airline_df[-train_ind, ]

#Using operational features as per EDA done earlier
logreg.fit <- glm(loss_factor ~ Load_Factor + Flight_Hours +
                   Route_Category + Aircraft_Capacity +
                   Season,
                   data = df_train,
                   family = binomial)

summary(logreg.fit)

pred_prob <- predict(logreg.fit, newdata = df_test, type = "response")
head(pred_prob)
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
head(pred_class)

table(Predicted = pred_class, Actual = df_test$loss_factor)

confusionMatrix(as.factor(pred_class),
                as.factor(df_test$loss_factor),
                positive = "1")


roc_obj <- roc(df_test$loss_factor, pred_prob)
auc_value <- auc(roc_obj)

plot(roc_obj,
     col = "blue",
     main = "ROC Curve for Route Loss Prediction")

legend("bottomright",
       legend = paste("AUC =", round(auc_value,3)))


```
Model 3


```{r}
#Defining the binary variable 0 - Profit 1 - Loss
airline_df$loss_factor <- ifelse(airline_df$route_classification == "Loss", 1, 0)

#Train and Test - Data splitting
set.seed(1)

train_ind <- sample(1:nrow(airline_df), size = nrow(airline_df)*0.8)

df_train <- airline_df[train_ind, ]
df_test <- airline_df[-train_ind, ]

#Using operational features as per EDA done earlier
logreg.fit <- glm(loss_factor ~ Load_Factor + Flight_Hours +
                   Route_Category + Aircraft_Type + Demand_Level +
                   Season,
                   data = df_train,
                   family = binomial)

summary(logreg.fit)

pred_prob <- predict(logreg.fit, newdata = df_test, type = "response")
head(pred_prob)
pred_class <- ifelse(pred_prob > 0.4, 1, 0)
head(pred_class)

table(Predicted = pred_class, Actual = df_test$loss_factor)

confusionMatrix(as.factor(pred_class),
                as.factor(df_test$loss_factor),
                positive = "1")


roc_obj <- roc(df_test$loss_factor, pred_prob)
auc_value <- auc(roc_obj)

png("roc_curve_model3.png")
plot(roc_obj,
     col = "blue",
     main = "ROC Curve for Route Loss Prediction")

legend("bottomright",
       legend = paste("AUC =", round(auc_value,3)))

dev.off()

```


```{r}
#Decision tree

library(tree)
library(caret)

# Create binary target variable - Loss or Profit
airline_df$loss_factor <- factor(
  ifelse(airline_df$route_classification == "Loss", "Loss", "Profit")
)

# Train-test split
set.seed(1)

train_indices <- sample(1:nrow(airline_df), nrow(airline_df) * 0.8)

train <- airline_df[train_indices, ]
test <- airline_df[-train_indices, ]

# Fit classification tree
airline.tree <- tree(
  loss_factor ~ Load_Factor + Flight_Hours + Route_Category +
    Aircraft_Type + Demand_Level + Season + Aircraft_Capacity,
  data = train
)

# View tree output
airline.tree

# Plot tree
plot(airline.tree)
text(airline.tree, pretty = 0, cex = 0.7)

# Training performance
summary(airline.tree)
```
```{r}
# Predict classes on test data
tree.pred <- predict(airline.tree, newdata = test, type = "class")

# Misclassification error rate
tree.error <- mean(tree.pred != test$loss_factor)
tree.error

# Confusion matrix
confusionMatrix(tree.pred, test$loss_factor, positive = "Loss")
```

```{r}
#Decision Tree with Pruning
set.seed(10)

# Cross-validation for pruning
cv.airline <- cv.tree(airline.tree, FUN = prune.misclass)

# Plot CV deviance against tree size
plot(cv.airline$size, cv.airline$dev, type = "b",
     xlab = "Tree Size",
     ylab = "Deviance",
     main = "Cross-validation for Tree Pruning")

# Best tree size
best.size <- cv.airline$size[which.min(cv.airline$dev)]
best.size

# Prune tree
pruned.airline.tree <- prune.misclass(airline.tree, best = best.size)

# Plot pruned tree
plot(pruned.airline.tree)
text(pruned.airline.tree, pretty = 0, cex = 0.7)
```
```{r}
# Predict using pruned tree
pruned.pred <- predict(pruned.airline.tree, newdata = test, type = "class")

# Misclassification error rate
pruned.error <- mean(pruned.pred != test$loss_factor)
pruned.error

# Confusion matrix
confusionMatrix(pruned.pred, test$loss_factor, positive = "Loss")
```


```{r}
#Randomforest
library(randomForest)

airline_df$loss_factor <- factor(
  ifelse(airline_df$route_classification == "Loss", "Loss", "Not_Loss")
)

set.seed(1)

train_indices <- sample(1:nrow(airline_df), nrow(airline_df) * 0.8)

df_train <- airline_df[train_indices, ]
df_test <- airline_df[-train_indices, ]

rf.airline <- randomForest(
  loss_factor ~ Load_Factor + Flight_Hours +
    Route_Category + Aircraft_Type +
    Demand_Level + Season,
  data = df_train,
  mtry = 3,
  ntree = 500,
  importance = TRUE
)

rf.airline
```

```{r}
rf.pred <- predict(rf.airline, newdata = df_test)

mean(rf.pred != df_test$route_classification)

importance(rf.airline)
varImpPlot(rf.airline)
```
```{r}
str(df_train$loss_num)
```

```{r}
#Boosting
library(gbm)

df_train$target_num <- ifelse(
  df_train$route_classification == "Loss", 1, 0
)
df_test$target_num <- ifelse(
  df_test$route_classification == "Loss", 1, 0
)
set.seed(1)

boost.airline <- gbm(
  target_num ~ Load_Factor + Flight_Hours +
    Route_Category + Aircraft_Type +
    Demand_Level + Season,
  data              = df_train,
  distribution      = "bernoulli",
  n.trees           = 1000,
  interaction.depth = 1,
  shrinkage         = 0.1,
  verbose           = FALSE
)

boost.airline

summary(boost.airline)

boost.pred <- ifelse(
  predict(boost.airline, newdata = df_test, n.trees = 1000) >= 0.5,
  1, 0
)

mean(boost.pred != df_test$target_num)

confusionMatrix(
  as.factor(boost.pred),
  as.factor(df_test$target_num),
  positive = "1"
)

#Explore different Parameters
set.seed(1)

boost.airline2 <- gbm(
  target_num ~ Load_Factor + Flight_Hours +
    Route_Category + Aircraft_Type +
    Demand_Level + Season,
  data              = df_train,
  distribution      = "bernoulli",
  n.trees           = 1000,
  interaction.depth = 2,
  shrinkage         = 0.05,
  verbose           = FALSE
)

boost.airline2

# Test model 2
boost.pred2 <- ifelse(
  predict(boost.airline2, newdata = df_test, n.trees = 1000) >= 0.5,
  1, 0
)

mean(boost.pred2 != df_test$target_num)

confusionMatrix(
  as.factor(boost.pred2),
  as.factor(df_test$target_num),
  positive = "1"
)

```
```{r}

boost.prob <- predict(boost.airline2, newdata = df_test, n.trees = 1000, type = "response")

boost.pred_04 <- ifelse(boost.prob >= 0.4, 1, 0)

confusionMatrix(
  as.factor(boost.pred_04),
  as.factor(df_test$target_num),
  positive = "1"
)

thresholds <- c(0.5, 0.45, 0.4, 0.35, 0.3)

for(t in thresholds){
  pred <- ifelse(boost.prob >= t, 1, 0)
  cm <- confusionMatrix(
    as.factor(pred),
    as.factor(df_test$target_num),
    positive = "1"
  )
  
  cat("\nThreshold:", t,
      "\nAccuracy:", round(cm$overall["Accuracy"], 4),
      "\nSensitivity:", round(cm$byClass["Sensitivity"], 4),
      "\nSpecificity:", round(cm$byClass["Specificity"], 4),
      "\n")
}

```

